Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2023-12-14 main.tex
:pencil: Finsh MIMO PC rate problem

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}

@@

		\begin{figure}[!t]
			\centering
			\subfloat[Without Direct Link\label{sm:pc_power_bond_nd}]{
				\resizebox{0.48\columnwidth}{!}{
					\input{assets/simulation/pc_power_bond_nd.tex}
				}
			}
			\subfloat[With Direct Link\label{sm:pc_power_bond_hd}]{
				\resizebox{0.48\columnwidth}{!}{
					\input{assets/simulation/pc_power_bond_hd.tex}
				}
			}
			\caption{Average channel power versus \gls{ris} group size $L$ for $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (8, 256, 4)$, $(\Lambda^\mathrm{D}, \Lambda^\mathrm{F}, \Lambda^\mathrm{B}) = (65, 54, 46) \unit{dB}$.}
			\label{sm:pc_power_bond}
		\end{figure}
		% Fig.~\ref{sm:pc_power_sx} shows that, apart from numerous reflecting elements, a sufficiently large group size can help
		% Fig.~\ref{sm:pc_power_sx} shows that increasing the group size can improve the channel power especially for a large \gls{ris}.
		Fig.~\ref{sm:pc_power_sx} shows that, apart from adding reflecting elements $N^\mathrm{S}$, increasing the group size $L$ also improves the channel power.
		% The behavior is more obvious for a large \gls{ris}.
		This behavior is more pronounced for a large \gls{ris}.
		For example, the gain of pairwise connection is 2.8\% for $N^\mathrm{S} = 16$ and 28\% for $N^\mathrm{S} = 256$.
		% \gls{bd} \gls{ris} provides a higher channel power than conventional \gls{ris}.
		% As $L$ increases from 1 to 2,
		It implies that the channel shaping capability of \gls{bd} \gls{ris} scales with group size $L$.

		Fig.~\ref{sm:pc_power_bond_hd} and \ref{sm:pc_power_bond_nd} compare the average channel power without and with direct link.
		% where the cascaded channel power available to passive \gls{ris} is $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% where the cascaded channel power available to passive \gls{ris} is $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% ``Cascaded'' means the maximum power of the cascaded channel, i.e., $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% maximum power of the cascaded channel, i.e., $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		``Cascaded'' means the \emph{power product} of the forward and backward channels.
		We observe that diagonal \gls{ris} wastes substantial cascaded power and struggles to align the direct-indirect eigenspace.
		When the direct link is absent, only 2.6\% of available power is utilized by diagonal \gls{ris} while 100\% power is recycled by fully-connected \gls{ris}.
		When the direct link is present, the proposed \gls{bd} \gls{ris} design can balance the direct-indirect and forward-backward eigenspace alignment for an optimal channel boost.
		It is worth noting that, when $L$ is sufficiently large, the composite channel power surpasses the power sum of direct and cascaded channels, thanks to the constructive \emph{amplitude superposition} of direct and cascaded channels.
		This again emphasizes the advantage of in-group connection of \gls{bd} \gls{ris}.
		% This is because the direct and indirect channels superpose
		% the proposed \gls{bd} \gls{ris} design can also be applied to conventional \gls{ris} to improve the channel power.
		% When the direct link is present, a small $L$ cannot align the direct-indirect eigenspace.

@@

		\eqref{op:pc_rate} is jointly non-convex and solved by \gls{ao}.
		For a given $\mathbf{\Theta}$, the optimal precoder is given by
		\begin{equation}
			\mathbf{W}^\star = \mathbf{V} {\mathbf{S}^\star}^{1/2},
		\end{equation}
		where $\mathbf{V}$ is right singular matrix of $\mathbf{H}$ and $\mathbf{S}^\star$ is a diagonal matrix of the water-filling power allocation.
		For a given $\mathbf{W}$, we update $\mathbf{\Theta}$ by \gls{rcg} method along the geodesics \cite{Abrudan2009}.
		\begin{remark}
			A geodesic refers to the shortest path between two points in a Riemannian manifold.
			% For unitary constraint \eqref{cs:pc_rate_unitary}, the
			Unitary constraint \eqref{cs:pc_rate_unitary} translates to a Stiefel manifold where the geodesics have simple expressions described by the exponential map \cite{Abrudan2008}.
			% Geodesics are the locally shortest paths in a Riemannian manifold.
			% A Geodesics are the locally shortest paths in a Riemannian manifold.
			% on Riemannian manifold are the shortest path between two points.
		\end{remark}
		For general optimization problems with block unitary constraint, the adapted \gls{rcg} method at iteration $r$ for block $g$ is summarized below, where $f^{(r)}$ is evaluated over $\bigl\{ \{\mathbf{\Theta}_{g'}^{(r+1)}\}_{g'=1}^{g-1}, \{\mathbf{\Theta}_{g'}^{(r)}\}_{g'=g}^{G} \bigr\}$.
		\begin{enumerate}
			\item Compute the Euclidean gradient
			\begin{equation}
				\nabla_g^\mathrm{E} f^{(r)} = \frac{\partial f^{(r)}}{\partial \mathbf{\Theta}_g^*};
				\label{eq:rcg_gradient_euclidean}
			\end{equation}
			\item Translate to the Riemannian gradient
			\begin{equation}
				\nabla_g^\mathrm{R} f^{(r)} = \nabla_g^\mathrm{E} f^{(r)} {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} - \mathbf{\Theta}_g^{(r)} {\nabla_g^\mathrm{E} f^{(r)}}^\mathsf{H};
				\label{eq:rcg_gradient_riemannian}
			\end{equation}
			\item Determine the weight factor
			\begin{equation}
				\gamma_g^{(r)} = \frac{\mathrm{tr}\Bigl(\bigl(\nabla_g^\mathrm{R} f^{(r)} - \nabla_g^\mathrm{R} f^{(r-1)}\bigr) \nabla_g^\mathrm{R} {f^{(r)}}^\mathsf{H}\Bigr)}{\mathrm{tr}\Bigl(\nabla_g^\mathrm{R} f^{(r-1)} \nabla_g^\mathrm{R} {f^{(r-1)}}^\mathsf{H}\Bigr)};
				\label{eq:rcg_weight}
			\end{equation}
			\item Compute the conjugate direction
			\begin{equation}
				\mathbf{D}_g^{(r)} = \nabla_g^\mathrm{R} f^{(r)} + \gamma_g^{(r)} \mathbf{D}_g^{(r-1)};
				\label{eq:rcg_direction}
			\end{equation}
			\item Determine the Armijo step size\footnote{To double the step size, simply square the argument instead of recomputing the matrix exponential, i.e., $\exp(2 \mu_g \mathbf{D}_g) = \exp^2(\mu_g \mathbf{D}_g)$.}
			\begin{equation}
				\mu_g^{(r)} = \arg\max_{\mu_g} f^{(r)}\Bigl(\exp\bigl(\mu_g \mathbf{D}_g^{(r)}\bigr) \mathbf{\Theta}_g^{(r)}\Bigr);
				\label{eq:rcg_step}
			\end{equation}
			\item Perform rotational update along local geodesics
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \exp\Bigl(\mu_g^{(r)} \mathbf{D}_g^{(r)}\Bigr) \mathbf{\Theta}_g^{(r)}.
				\label{eq:rcg_update}
			\end{equation}
		\end{enumerate}

@@

		The complex derivative of \eqref{ob:pc_rate} w.r.t. \gls{ris} block $g$ is
		\begin{equation}
			\frac{\partial R}{\partial \mathbf{\Theta}_g^*} = \frac{1}{\sigma_n^2} {\mathbf{H}_g^\mathrm{B}}^\mathsf{H} \mathbf{HW} \biggl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\sigma_n^2}\biggr)^{-1} \mathbf{W}^\mathsf{H} {\mathbf{H}_g^\mathrm{F}}^\mathsf{H}.
			\label{eq:pc_rate_gradient_ris}
		\end{equation}
		The adapted \gls{rcg} method for the \gls{ris} rate maximization subproblem is summarized in Algorithm~\ref{ag:pc_rate_ris}.

		\setalgorithmcaptionfont{\small}
		\begin{algorithm}[!t]
			\small
			\caption{\gls{rcg} Method for \gls{ris} \gls{mimo}-\gls{pc} Rate Maximization}
			\label{ag:pc_rate_ris}
			\begin{algorithmic}[1]
				\Require $\mathbf{H}^\mathrm{D}$, $\mathbf{H}^\mathrm{F}$, $\mathbf{H}^\mathrm{B}$, $\mathbf{W}$, $L$, $\sigma_n^2$
				\Ensure $\mathbf{\Theta}^\star$
				% \State Initialize $r \gets 0$, $\mathbf{\Theta}^{(0)}$
				\State $r \gets 0$, $\mathbf{\Theta}^{(0)}$
				\Repeat
					% \State Update $r \gets r+1$
					\State $r \gets r+1$
					\For {$g \gets 1$ to $G$}
						% \State Get $\mathbf{\Theta}_g^{(r)}$ by \eqref{eq:pc_rate_gradient_ris}, \eqref{eq:rcg_gradient_riemannian} -- \eqref{eq:rcg_update}
						\State $\mathbf{\Theta}_g^{(r)} \gets$ \eqref{eq:pc_rate_gradient_ris}, \eqref{eq:rcg_gradient_riemannian}--\eqref{eq:rcg_update}
					\EndFor
				\Until $\lvert R^{(r)} - R^{(r-1)} \rvert / R^{(r-1)} \le \epsilon$
			\end{algorithmic}
		\end{algorithm}
	\end{subsection}
\end{section}

>>>>

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}

@@

		\begin{figure}[!t]
			\centering
			\subfloat[Without Direct Link\label{sm:pc_power_bond_nd}]{
				\resizebox{0.48\columnwidth}{!}{
					\input{assets/simulation/pc_power_bond_nd.tex}
				}
			}
			\subfloat[With Direct Link\label{sm:pc_power_bond_hd}]{
				\resizebox{0.48\columnwidth}{!}{
					\input{assets/simulation/pc_power_bond_hd.tex}
				}
			}
			\caption{Average channel power versus \gls{ris} group size $L$ for $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (8, 256, 4)$, $(\Lambda^\mathrm{D}, \Lambda^\mathrm{F}, \Lambda^\mathrm{B}) = (65, 54, 46) \unit{dB}$.}
			\label{sm:pc_power_bond}
		\end{figure}
		% Fig.~\ref{sm:pc_power_sx} shows that, apart from numerous reflecting elements, a sufficiently large group size can help
		% Fig.~\ref{sm:pc_power_sx} shows that increasing the group size can improve the channel power especially for a large \gls{ris}.
		Fig.~\ref{sm:pc_power_sx} shows that, apart from adding reflecting elements $N^\mathrm{S}$, increasing the group size $L$ also improves the channel power.
		% The behavior is more obvious for a large \gls{ris}.
		This behavior is more pronounced for a large \gls{ris}.
		For example, the gain of pairwise connection is \qty{2.8}{\percent} for $N^\mathrm{S} = 16$ and \qty{28}{\percent} for $N^\mathrm{S} = 256$.
		% \gls{bd} \gls{ris} provides a higher channel power than conventional \gls{ris}.
		% As $L$ increases from 1 to 2,
		It implies that the channel shaping capability of \gls{bd} \gls{ris} scales with group size $L$.

		Fig.~\ref{sm:pc_power_bond_hd} and \ref{sm:pc_power_bond_nd} compare the average channel power without and with direct link.
		% where the cascaded channel power available to passive \gls{ris} is $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% where the cascaded channel power available to passive \gls{ris} is $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% ``Cascaded'' means the maximum power of the cascaded channel, i.e., $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% maximum power of the cascaded channel, i.e., $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		``Cascaded'' means the \emph{power product} of the forward and backward channels.
		We observe that diagonal \gls{ris} wastes substantial cascaded power and struggles to align the direct-indirect eigenspace.
		When the direct link is absent, only \qty{2.6}{\percent} of available power is utilized by diagonal \gls{ris} while \qty{100}{\percent} power is recycled by fully-connected \gls{ris}.
		When the direct link is present, the proposed \gls{bd} \gls{ris} design can balance the direct-indirect and forward-backward eigenspace alignment for an optimal channel boost.
		It is worth noting that, when $L$ is sufficiently large, the composite channel power surpasses the power sum of direct and cascaded channels, thanks to the constructive \emph{amplitude superposition} of direct and cascaded channels.
		This again emphasizes the advantage of in-group connection of \gls{bd} \gls{ris}.
		% This is because the direct and indirect channels superpose
		% the proposed \gls{bd} \gls{ris} design can also be applied to conventional \gls{ris} to improve the channel power.
		% When the direct link is present, a small $L$ cannot align the direct-indirect eigenspace.

@@

		\eqref{op:pc_rate} is jointly non-convex and solved by \gls{ao}.
		For a given $\mathbf{\Theta}$, the optimal precoder is given by
		\begin{equation}
			\mathbf{W}^\star = \mathbf{V} {\mathbf{S}^\star}^{1/2},
		\end{equation}
		where $\mathbf{V}$ is right singular matrix of $\mathbf{H}$ and $\mathbf{S}^\star$ is a diagonal matrix of the water-filling power allocation.
		For a given $\mathbf{W}$, we update $\mathbf{\Theta}$ by \gls{rcg} method along the geodesics \cite{Abrudan2009}.
		\begin{remark}
			A geodesic refers to the shortest path between two points in a Riemannian manifold.
			% For unitary constraint \eqref{cs:pc_rate_unitary}, the
			Unitary constraint \eqref{cs:pc_rate_unitary} translates to a Stiefel manifold where the geodesics have simple expressions described by the exponential map \cite{Abrudan2008}.
			% Geodesics are the locally shortest paths in a Riemannian manifold.
			% A Geodesics are the locally shortest paths in a Riemannian manifold.
			% on Riemannian manifold are the shortest path between two points.
		\end{remark}
		For general optimization problems with block unitary constraint, the adapted \gls{rcg} method at iteration $r$ for block $g$ is summarized below, where $f\bigl(\mathbf{\Theta}_g^{(r)}\bigr)$ is the objective function evaluated over $\bigl\{ \{\mathbf{\Theta}_{g'}^{(r+1)}\}_{g'<g}, \{\mathbf{\Theta}_{g'}^{(r)}\}_{g'>g} \bigr\}$.
		\begin{enumerate}
			\item Compute the Euclidean gradient
			\begin{equation}
				{\nabla_g^\mathrm{E}}^{(r)} = \frac{\partial f\bigl(\mathbf{\Theta}_g^{(r)}\bigr)}{\partial \mathbf{\Theta}_g^*};
				\label{eq:rcg_gradient_euclidean}
			\end{equation}
			\item Translate to the Riemannian gradient
			\begin{equation}
				{\nabla_g^\mathrm{R}}^{(r)} = {\nabla_g^\mathrm{E}}^{(r)} {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} - \mathbf{\Theta}_g^{(r)} {{\nabla_g^\mathrm{E}}^{(r)}}^\mathsf{H};
				\label{eq:rcg_gradient_riemannian}
			\end{equation}
			\item Determine the weight factor
			\begin{equation}
				\gamma_g^{(r)} = \frac{\mathrm{tr}\Bigl(\bigl({\nabla_g^\mathrm{R}}^{(r)} - {\nabla_g^\mathrm{R}}^{(r-1)}\bigr) {{\nabla_g^\mathrm{R}}^{(r)}}^\mathsf{H}\Bigr)}{\mathrm{tr}\Bigl({\nabla_g^\mathrm{R}}^{(r-1)} {{\nabla_g^\mathrm{R}}^{(r-1)}}^\mathsf{H}\Bigr)};
				\label{eq:rcg_weight}
			\end{equation}
			\item Compute the conjugate direction
			\begin{equation}
				\mathbf{D}_g^{(r)} = {\nabla_g^\mathrm{R}}^{(r)} + \gamma_g^{(r)} \mathbf{D}_g^{(r-1)};
				\label{eq:rcg_direction}
			\end{equation}
			\item Determine the Armijo step size\footnote{To double the step size, simply square the argument instead of recomputing the matrix exponential, i.e., $\exp(2 \mu_g \mathbf{D}_g) = \exp^2(\mu_g \mathbf{D}_g)$.}
			\begin{equation}
				\mu_g^{(r)} = \arg\max_{\mu_g} f\Bigl(\exp\bigl(\mu_g \mathbf{D}_g^{(r)}\bigr) \mathbf{\Theta}_g^{(r)}\Bigr);
				\label{eq:rcg_step}
			\end{equation}
			\item Perform rotational update along local geodesics
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \exp\Bigl(\mu_g^{(r)} \mathbf{D}_g^{(r)}\Bigr) \mathbf{\Theta}_g^{(r)}.
				\label{eq:rcg_update}
			\end{equation}
		\end{enumerate}

@@

		The complex derivative of \eqref{ob:pc_rate} w.r.t. \gls{ris} block $g$ is
		\begin{equation}
			\frac{\partial R}{\partial \mathbf{\Theta}_g^*} = \frac{1}{\sigma_n^2} {\mathbf{H}_g^\mathrm{B}}^\mathsf{H} \mathbf{HW} \biggl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\sigma_n^2}\biggr)^{-1} \mathbf{W}^\mathsf{H} {\mathbf{H}_g^\mathrm{F}}^\mathsf{H}.
			\label{eq:pc_rate_gradient_ris}
		\end{equation}
		Algorithm~\ref{ag:pc_rate_ris} summarizes the adapted \gls{rcg} method for the \gls{ris} rate maximization subproblem.

		\setalgorithmcaptionfont{\small}
		\begin{algorithm}[!t]
			\small
			\caption{\gls{rcg} Method for \gls{ris} \gls{mimo}-\gls{pc} Rate Maximization}
			\label{ag:pc_rate_ris}
			\begin{algorithmic}[1]
				\Require $\mathbf{H}^\mathrm{D}$, $\mathbf{H}^\mathrm{F}$, $\mathbf{H}^\mathrm{B}$, $\mathbf{W}$, $L$, $\sigma_n^2$
				\Ensure $\mathbf{\Theta}^\star$
				\State $r \gets 0$, $\mathbf{\Theta}^{(0)}$
				\Repeat
					\State $r \gets r+1$
					\For {$g \gets 1$ to $G$}
						\State $\mathbf{\Theta}_g^{(r)} \gets$ \eqref{eq:pc_rate_gradient_ris}, \eqref{eq:rcg_gradient_riemannian}--\eqref{eq:rcg_update}
					\EndFor
				\Until $\lvert R^{(r)} - R^{(r-1)} \rvert / R^{(r-1)} \le \epsilon$
			\end{algorithmic}
		\end{algorithm}

		\begin{figure}[!t]
			\centering
			\resizebox{0.65\columnwidth}{!}{
				\input{assets/simulation/pc_rate_sx.tex}
			}
			\caption{Average achievable rate versus \gls{ris} elements $N^\mathrm{S}$ and group size $L$ for $(N^\mathrm{T}, N^\mathrm{R}) = (8, 4)$, $(\Lambda^\mathrm{D}, \Lambda^\mathrm{F}, \Lambda^\mathrm{B}) = (65, 54, 46) \unit{dB}$.}
			\label{sm:pc_rate_sx}
		\end{figure}

		Fig.~\ref{sm:pc_rate_sx} illustrates how \gls{ris} configuration influences the \gls{mimo} \gls{pc} achievable rate.
		To ensure a \qty{20}{bit/s/Hz} transmission, an \gls{snr} of \qty{13.5}{\dB} is required for a 8T4R system.
		This value decreases to \qty{12.5}{\dB} (resp. \qty{8}{\dB}) when 32- (resp. 256-) element diagonal \gls{ris} is present.
		If tetrads can be formed in \gls{bd} \gls{ris}, the \gls{snr} can be reduced by another \qty{20}{\percent} (resp. \qty{44}{\percent}).
		Further increase in $L$ yields a marginal gain and incurs quadratic number of connections.
		Thus, dyadic or tetradic \gls{bd} \gls{ris} usually strike a good balance between performance and complexity.

	\end{subsection}
\end{section}
